### 2020-02-24 thesis/paper.tex
edits

\part{Theory}
\chapter{Methods}\label{cha:methods}
The newly created tool uses a large amount of existing methods which are described in this chapter. Here, I aim to give an overview of what exact methods were used, the theory behind them and in which scenarios they can be/are applied. For a more general overview of some of the topics relating the methods used, see the related work of this paper in Chapter \ref{cha:related_work}.

\section{Dimensionality Reduction}\label{sec:dim_reduction}
Dimensionality reduction are able to reduce the number of dimensions in a data-set while trying to maintain as much information as possible. Reducing the number of dimensions is often needed for reducing the computational complexity of clustering and improving the accuracy of the result, but also to visualize the data in a lower-dimensional space. This visualization is then often easier to understand for a user than if all dimensions are shown at once.
\subsection{Principal Component Analysis}
 Principal Component Analysis (PCA) \cite{pca} aims to transform correlated variables into uncorrelated ones thereby omitting less relevant dimensions. Generally, Principal Components (PCs) are linear combinations of the original variables and can be computed through the eigenvectors $V$ of the covariance matrix $C$. To use these PCs for dimensionality reduction the $k$ onces corresponding to the largest eigenvalues of the matrix are extracted into a matrix $V_k$ and used as a basis to transform the data. This is done by calculating $data_{red.}=V_k^T*data_{center}$, whereby the original data must be centered first.

\subsection{t-Distributed Stochastic Neighbor Embedding}
The dimensionality recution method t-Distributed Stochastic Neighbor Embedding (t-SNE) \cite{Maaten2008VisualizingDU} works by converting the Euclidean distances between points into conditional probabilities that represent similarities. This similarity represents the probability that a point would be picked as the neighbor of another point if neighbors were picked proportionally to their probability density under a Student-t distribution centered at a given point. To lay out the points, an initial randomly sampled solution is created and improved on iteratively with gradient decent, such that the similarities are best represented by the points in low-dimensional space. As inputs the number of dimensions $k$ and the perplexity and some optimization parameters are needed. The number of resulting dimensions is defined by $k$ and the optimization parameters can be used to improve the runtime of the algorithm. The perplexity parameter can be seen as a smoothing parameter defining the effective number of neighbors for each point. This parameter is usually also fairly robust \cite{Maaten2008VisualizingDU}, good choices in practice are between $5$ and $50$.

\subsection{Multi-Dimensional Scaling}
Multi-Dimensional Scaling (MDS) \cite{mds} uses a distance matrix (often called proximity data) and a number of target dimensions $k$ as input. It then trys to lay out the point in a $k$-dimensional space such that the distances between them in that space are as close as possible to the true distance. When this method is used with a distance matrix created with Euclidean distances, the result will be equivalent to running PCA on the data itself.

For the calculation of the solution MDS builds $n$ $k$-dimensional hypervectors with unknown parameters and aims to minimize the stress value by applying gradient decent. The stress of a configuration is defined as the residual sum of squares for the distances, i.e. the sum of squared errors between the actual distances and the distances in the $k$-dimensional space.

\section{Clustering}
The clustering algorithms implemented were used to create base clusterings for meta-clustering. Besides this, any base clustering result may be chosen by the user as the overall result. The algorithms described here are the ones used in the created tool to generate base clusterings, except for OPTICS which was used for meta-clustering. For additional information on the choice of OPTICS for the meta-level clustering, see Section \ref{sec:sel_alg_param}.

@@

The main idea of k-Means type clustering algorithms is to start with centroids which are points in the data-space representing the centers of the clusters. Each point is assigned to the centroid closest to it according to Euclidean distance, defining the clusters. To improve the result the centroid is updated to the center of its corresponding points iteratively until no more changes happen or some convergence criterion is met. This update procedure does not guarantee finding the optimum though, even if no more changes for the centroids can be observed. How the initial centroids are chosen may also impact the quality of the result greatly. To find a good initialization, Arthur and Vassilvitskii \cite{10.5555/1283383.1283494} proposed k-Means++ whereby the centroids are picked such that they are far apart with high probability. They showed that this kind of initialization makes the overall algorithm $O(log(k))$ competitive with the optimal clustering \cite{10.5555/1283383.1283494}.

@@

\subsection{Expectation Maximization}
The Expectation Maximization (EM) algorithm \cite{10.2307/2984875} aims to iteratively find estimates of parameters of a model such that they represent the maximum likelihood of those. To do this, the algorithm alternates between two steps, those being the expectation step and the maximization step. In the expectation step the expectation of the log-likelihood is defined as a function given the current set of parameters. The maximization step then computes the new set of parameters such that the expected log-likelihood function is maximized.

This can be applied for clustering by defining the model as a Gaussian mixture and running this EM algorithm such that the best model given the parameter $k$ is found. A benefit of this can be, that Gaussian clusters with different spread can more easily be found than with k-Means.

\subsection{DBSCAN}
Density-Based Spatial Clustering of Application with noise (DBSCAN) \cite{10.5555/3001460.3001507}, is an algorithm that unlike k-Means is not based on partitioning the data directly but on searching for dense regions in the data. For the calculation two parameters are needed, Epsilon ($\epsilon$) and a minimum number of objects ($\mu$). The parameter $\epsilon$ defines how large a region may be considered for neighbor search and $\mu$ defines how many objects must be within that region for it to be considered dense. The following explanations are needed for full understanding of the algorithm:
\newline

When defining the neighborhood within the epsilon range for an object $p$ as $N_\epsilon(p)$ the algorithm considers an object $p$ as a core object if $|N_\epsilon(p)|\geq \mu$.

@@

For two objects $p$ and $q$ they are density-reachable if there exists a path of objects between them for which each pair of objects is directly density-reachable.

@@

Also the relationship between $\epsilon$ and $\mu$ is interesting to observe in practice. Often when trying different parameter sets for DBSCAN is suffices to fix one parameter to a somewhat sensible value while only sampling the other. If one for example knows that any region with less then $20$ ``close'' points should not be considered a cluster it is possible to fix $\mu$ at $20$ and only sample $\epsilon$ to find a satisfactory result.

\subsection{OPTICS}\label{sec:opticstheory}
Ordering Points to Identify the Clustering Structure or OPTICS \cite{10.1145/304181.304187} is an extension of DBSCAN whereby DBSCANs Epsilon $\epsilon$ parameter is used in a different manner. It also has a parameter MinPts which has a similar meaning ad the $\mu$ parameter of DBSCAN. The idea is that different distance parameters are used in each step such that density-based clusters with different densities can be found simultaneously. To achieve this the  $\epsilon$ parameter only defines the maximum distance for which points can be considered reachable and the actual reachability distance for two objects is computed as:
$$reachability\text{-}distance_{\epsilon,MinPts}(p,o)=$$
$$
\begin{cases}
    \text{UNDEFINED},& \text{if } |N_\epsilon(p)|< \text{MinPts}\\
    max(core\text{-}distance(o),distance(o,p)),              & \text{otherwise}
\end{cases}
$$
whereby:
$$core\text{-}distance_{\epsilon,MinPts}(p)=$$
$$
\begin{cases}
    \text{UNDEFINED},& \text{if } |N_\epsilon(p)|< \text{MinPts}\\
    MinPts\text{-}distance(p),              & \text{otherwise}
\end{cases}
$$

@@

\section{Consensus Functions}
Consensus functions have the goal of merging multiple different clustering results together such that the overall result represents a consensus between the base clusterings. In this work I focused on two methods that aim to achieve this.
\subsection{Co-association Matrix based Consensus}
Co-association Matrix based consensus functions \cite{Monti2003} first create a co-association matrix that indicates how often two points $p$ and $q$ appear in in a common cluster, i.e. if $p$ and $q$ are clustered together $70\%$ of the time the corresponding entry of the matrix is $0.7$. To extract a consensus result from this matrix different methods can be applied. One possibility is to choose a clustering algorithm that takes this matrix as an input, converts it to a dissimilarity/distance measure and performs clustering with it. Another possibility is to use linking of sets strategies, which is also done for the implementation in the tool. This has the benefit of being trivial to implement and not imposing a cluster structure, as would for example be the case when using k-Means.The possible linking strategies are single-link, using the similarity of the two points of the sets where it is maximal, average-link, using the average similarity between all point pairs, or maximum link, whereby the smallest similarity is used. With each of these strategies the pair of sets for which this linking value is maximal (i.e. they are most similar) is merged. To stop the merging process and obtain the final result, either a threshold value can be set for the link value upon which no more merging should be performed or a number of target clusters $k$ can be defined stopping when only $k$ sets remain. Another option for defining when to stop merging is the Lifetime Criterion. For the implementation of the tool per default the average-link strategy is chosen with either defining $k$ or using this Lifetime Criterion.

\subsubsection{Lifetime Criterion}
The lifetime of a group of sets is defined as the difference between the linking values for the previous merge and the next merge. These linking values are the maximum link value according to some strategy, where the next merge of two sets would occur. The Lifetime Criterion claims that a good choice for a cut in a hierarchical clustering tree is the one for which the lifetime is maximal \cite{YANG201735}. This estimate similarly to the elbow-method for other clustering algorithms aims to find a good trade-off between the number of clusters and the accuracy decrease due to increasing/decreasing the cluster number. To use this criterion, either two runs of merging (only keeping the linking values) need to be performed or the full merge-tree must be saved. If the whole tree is saved the result can be obtained by cutting the tree, otherwise the group of sets needs to be re-computed up to the corresponding level.

\subsection{DICLENS}
Divisive Clustering Ensemble with Automatic Cluster Number (DICLENS) \cite{DICLENS} organizes the clusterings in a graph with the edges representing inter-cluster similarities. Within this graph a minimum spanning tree is searched for with Prim's algorithm \cite{prim} and the ideal number of clusters $k$ is searched for. This $k$ is found by iteratively removing the edge with minimum weight and regarding the connected components as meta-clusters. For this meta-clustering majority voting is performed, computing a final result for which the intracluster similarity (ICS) and intercluster similarity (ECS) are used to measure the quality. This is repeated until all edges of the minimum spanning tree are removed and the result for which ICS-ECS is largest is chosen.

Mimaroglu and Aksehirli \cite{DICLENS} also provide a Java implementation of their method. It also includes a visual interface for simpler understanding. This implementation was also used to implement their method in this tool.

@@

\subsection{Kuhn’s Hungarian method}
Kuhn’s Hungarian method \cite{Kuhn2010} can be used to solve the mathematical assignment problem for the two-dimensional case. A simple example for the assignment problem is to think of $n$ jobs and $n$ workers in a factory that each demand a different payment for a specific job. Now the task is to find the optimal assignment whereby all jobs are performed, all workers are assigned a job and the total cost for all workers is minimal. The assignment problem also allows for reformulation such that the target value is maximized instead of minimized whenever this better fits the problem and it can be extended to more than two dimensions. In the aforementioned example an additional dimension could for example be the tools used, i.e. a worker demands a different payment depending on which task he has to perform with which tool being provided. Also, as an extension to make this more general each dimension may not necessarily need to have the same number of elements. If there are for example more workers than jobs, some workers will not be assigned a job.

image %%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The resulting function can then be used to analyze from which distributions the data-points may have been drawn which may also prove useful when analyzing clustering solutions.

@@

At first, the user tried sampling k-means with $k$ in the range $2$ to $20$ creating $10$ solutions each. Doing this, he looked at which values for $k$ resulted in outlying/non-robust results. When filtering for solutions with $\leq10$ clusters, it was easy to see that the results with more than $10$ clusters produce poor solutions. The effect of which clusterings are filtered out this way can be seen in Figure \ref{fig:user_qcm_optics}.

@@

To start of, the user tried creating clustering solutions using DBSCAN, but in both the reachability plot and the heat map it could be seen that no robust solutions except nearly trivial ones could be found. In most cases the points were either nearly all assigned to one cluster or different small clusters each. With this information he deducted, that the algorithm is probably not fit for finding a good solution and chose to use k-means. There, as he did not know how many clusters are in the data-set, he sampled $k$ from $2$ to $50$ and found again that solutions with $k>20$ seemed poor. He decided on this value, just like in the first example, by finding where solutions begin to differ strongly. The resulting reachability plot is shown in Figure \ref{fig:user_wifi_optics}

>>>>

\part{Theory}
\chapter{Methods}\label{cha:methods}
The newly created tool uses a large amount of existing methods which are described in this chapter. Here, I aim to give an overview of what exact methods were used, the theory behind them and in which scenarios they can be/are applied. For a more general overview of some of the topics relating the methods used, refer to the related work of this paper in Chapter \ref{cha:related_work}.

\section{Dimensionality Reduction}\label{sec:dim_reduction}
Dimensionality reduction techniques are able to reduce the number of dimensions in a data-set while trying to maintain as much information as possible. Reducing the number of dimensions is often needed for reducing the computational complexity of clustering and improving the accuracy of the result, but also to visualize the data in a lower-dimensional space. This visualization is then often easier to understand for a user than if all dimensions are shown at once.
\subsection{Principal Component Analysis}
 Principal Component Analysis (PCA) \cite{pca} aims to transform correlated variables into uncorrelated ones thereby omitting less relevant dimensions. Generally, Principal Components (PCs) are linear combinations of the original variables and can be computed through the eigenvectors $V$ of the covariance matrix $C$. To use these PCs for dimensionality reduction, the $k$ onces corresponding to the $k$ largest eigenvalues of the matrix are extracted into a matrix $V_k$ and used as a basis to transform the data. This is done by calculating $data_{reduced}=V_k^T*data_{center}$, whereby the original data must be centered first.

\subsection{t-Distributed Stochastic Neighbor Embedding}
The dimensionality recution method t-Distributed Stochastic Neighbor Embedding (t-SNE) \cite{Maaten2008VisualizingDU} works by converting the Euclidean distances between points into conditional probabilities that represent similarities. This similarity represents the probability that a point would be picked as the neighbor of another point if neighbors were picked proportionally to their probability density under a Student-t distribution centered at a given point. To lay out the points, an initial randomly sampled solution is created and improved on iteratively with gradient decent, such that the similarities are best represented by the points in low-dimensional space. As inputs for the method, the number of dimensions $k$, the perplexity and some optimization parameters are needed. The number of resulting dimensions is defined by $k$ and the optimization parameters can be used to improve the runtime of the algorithm. The perplexity parameter can be seen as a smoothing parameter defining the effective number of neighbors for each point. This parameter is usually also fairly robust \cite{Maaten2008VisualizingDU}, good choices in practice are values between $5$ and $50$.

\subsection{Multi-Dimensional Scaling}
Multi-Dimensional Scaling (MDS) \cite{mds} uses a distance matrix (often called proximity data) and a number of target dimensions $k$ as input. It then trys to lay out the points in a $k$-dimensional space such that the distance between each pair in that space is as close as possible to its true distance. When this method is used with a distance matrix created with Euclidean distances, the result will be equivalent to simply running the PCA method on the data itself.

For the calculation of the solution, MDS builds $n$ $k$-dimensional hypervectors with unknown parameters and aims to minimize the stress value by applying gradient decent. The stress of a configuration is defined as the residual sum of squares for the distances, i.e. the sum of squared errors between the actual distances and the distances in the $k$-dimensional space.

\section{Clustering}
The clustering algorithms implemented for the tool were used to create base clusterings for meta-clustering. Besides this, any base clustering result may be chosen by the user as the overall result. In contrast to the other algorithms, OPTICS was not used for the generation of base clusterings but to compute the meta-clustering. For additional information on the choice of OPTICS for meta-level clustering, see Section \ref{sec:sel_alg_param}.

@@

The main idea of k-Means type clustering algorithms is to start with centroids which are points in the data-space representing the centers of clusters. Each point is therby assigned to the centroid closest to it according to Euclidean distance, defining the these clusters. To improve the result, the centroid is updated to the center of its corresponding points iteratively. This is repeated until no more changes happen or some convergence criterion is met. The update procedure of each iteration does not guarantee finding the optimum though, even if no more changes for the centroids can be observed. How the initial centroids are chosen may also impact the quality of the result greatly. To find a good initialization, Arthur and Vassilvitskii \cite{10.5555/1283383.1283494} proposed k-Means++ whereby the centroids are picked such that they are far apart with high probability. They showed that this kind of initialization makes the overall algorithm $O(log(k))$ competitive with the optimal clustering \cite{10.5555/1283383.1283494}. Still, the usual choice is to initialize them by picking random data points.

@@

\subsection{Expectation Maximization}
The Expectation Maximization (EM) algorithm \cite{10.2307/2984875} aims to iteratively find estimates of parameters of a model such that they are assigned the most likely values. To do this, the algorithm alternates between two steps, those being the expectation step and the maximization step. In the expectation step the expectation of the log-likelihood is defined as a function given the current set of parameters. The maximization step then computes the new set of parameters such that the expected log-likelihood function is maximized.

This can be applied for clustering by defining the model as a Gaussian mixture and running this EM algorithm such that the best model, given the parameter $k$, is found. A benefit of this can be that Gaussian clusters with different distribution parameters can more easily be found than with k-Means.

\subsection{DBSCAN}
Density-Based Spatial Clustering of Application with Noise (DBSCAN) \cite{10.5555/3001460.3001507}, is an algorithm that unlike k-Means is not based on partitioning the data directly but on searching for dense regions in the data. For the calculation, two parameters are needed, namely epsilon ($\epsilon$) and a minimum number of objects ($\mu$ or MinPts). The parameter $\epsilon$ defines how large a region may be considered for neighborhood search and $\mu$ defines how many objects must be within that region for it to be considered dense. The following explanations are needed for full understanding of the algorithm:
\newline

When defining the neighborhood within the epsilon range for an object $p$ as $N_\epsilon(p)$, the algorithm considers an object $p$ as a core object if $|N_\epsilon(p)|\geq \mu$.

@@

For two objects $p$ and $q$, they are density-reachable if there exists a path of objects between them for which each pair of objects is directly density-reachable.

@@

Also, the relationship between $\epsilon$ and $\mu$ is interesting to observe in practice. Often, when trying different parameter sets for DBSCAN is suffices to fix one parameter to a somewhat sensible value while only sampling the other. If one for example knows that any region with less then $20$ ``close'' points should not be considered a cluster, it is possible to fix $\mu$ at $20$ and only sample $\epsilon$ to find a satisfactory result.

\subsection{OPTICS}\label{sec:opticstheory}
Ordering Points To Identify the Clustering Structure or OPTICS \cite{10.1145/304181.304187} is an extension of DBSCAN whereby DBSCAN's Epsilon $\epsilon$ parameter is used in a different manner. It also has a parameter MinPts which has a similar meaning as the $\mu$ parameter of DBSCAN. The idea is that different distance parameters are used in each step such that density-based clusters with different densities can be found simultaneously. To achieve this the  $\epsilon$ parameter only defines the maximum distance for which points can be considered reachable and the actual reachability distance for two objects is computed as:
$$reachability\text{-}distance_{\epsilon,MinPts}(p,o)=$$
$$
\begin{cases}
    \text{UNDEFINED},& \text{if } |N_\epsilon(p)|< \text{MinPts}\\
    max(core\text{-}distance(o),distance(o,p)),              & \text{otherwise}
\end{cases}
$$
whereby:
$$core\text{-}distance_{\epsilon,MinPts}(p)=$$
$$
\begin{cases}
    \text{UNDEFINED},& \text{if } |N_\epsilon(p)|< \text{MinPts}\\
    MinPts\text{-}distance(p),              & \text{otherwise}
\end{cases}
$$

@@

\section{Consensus Functions}
Consensus functions have the goal of merging multiple different clustering solutions together such that the overall result represents a consensus between all base clusterings. In this work I focused on two methods that aim to achieve this.
\subsection{Co-association Matrix based Consensus}
Co-association Matrix based consensus functions \cite{Monti2003} first create a co-association matrix that indicates how often two points $p$ and $q$ appear in in a common cluster, i.e. if $p$ and $q$ are clustered together $70\%$ of the time the corresponding entry of the matrix is $0.7$. To extract a consensus result from this matrix different methods can be applied. One possibility is to choose a clustering algorithm that takes this matrix as an input, converts it to a dissimilarity/distance measure and performs clustering with it. Another possibility is to use linking of sets strategies, which is also done for the implementation in the tool. This has the benefit of being trivial to implement and not imposing a cluster structure, as would for example be the case when using k-Means. The possible linking strategies are single-link, using the similarity of the two points of the sets where it is maximal, average-link, using the average similarity between all point pairs, or maximum link, whereby the smallest similarity is used. With each of these strategies the pair of sets for which this linking value is maximal (i.e. they are most similar) is merged. To stop the merging process and obtain the final result either a threshold value can be set for the link value upon which no more merging should be performed or a number of target clusters $k$ can be defined, stopping when only $k$ sets remain. Another option for defining when to stop merging is the Lifetime Criterion. For the implementation of the tool the average-link strategy is chosen per default with either defining $k$ or using this Lifetime Criterion such that is estimates $k$ itself.

\subsubsection{Lifetime Criterion}
The lifetime of a group of sets is defined as the difference between the linking values for the previous merge and the next merge. These linking values are the maximum link value according to some strategy, where the next merge of two sets would occur. The Lifetime Criterion claims that a good choice for a cut in a hierarchical clustering tree is the one for which the lifetime is maximal \cite{YANG201735}. This estimate, similarly to the elbow-method for other clustering algorithms, aims to find a good trade-off between the number of clusters and the accuracy decrease due to increasing/decreasing the cluster number. To use this criterion either two runs of merging (only keeping the linking values) need to be performed or the full merge-tree must be saved. If the whole tree is saved, the result can be obtained by cutting the tree, otherwise the grouping of sets needs to be re-computed up to the corresponding level.

\subsection{DICLENS}
Divisive Clustering Ensemble with Automatic Cluster Number (DICLENS) \cite{DICLENS} organizes the clusterings in a graph with the edges representing inter-cluster similarities. Within this graph a minimum spanning tree is searched for with Prim's algorithm \cite{prim} and the ideal number of clusters $k$ is searched for. This $k$ is found by iteratively removing the edge with minimum weight and regarding the connected components as meta-clusters. Next, majority voting is performed with the given meta-clustering, computing a final result for which the intracluster similarity (ICS) and intercluster similarity (ECS) are used as indicators of the quality. This is repeated until all edges of the minimum spanning tree are removed and the result for which ICS-ECS is largest is chosen.

Mimaroglu and Aksehirli \cite{DICLENS} also provide a Java implementation of their method, it also includes a visual interface for simpler understanding. This implementation was also used to implement their method my this tool.

@@

\subsection{Kuhn’s Hungarian method}
Kuhn’s Hungarian method \cite{Kuhn2010} can be used to solve the mathematical assignment problem for the two-dimensional case. A simple example for the assignment problem is to think of $n$ jobs and $n$ workers in a factory that each demand a different payment for each specific job. Now the task is to find the optimal assignment whereby all jobs are performed, all workers are assigned a job and the total cost for all workers is minimal. The assignment problem also allows for reformulation such that the target value is maximized instead of minimized, whenever this better fits the problem and it can be extended to more than two dimensions. In the aforementioned example an additional dimension could for example be the tools used, i.e. a worker demands a different payment depending on which task he has to perform with which tool being provided. Also, as an extension to make this more general each dimension may not necessarily need to have the same number of elements. If there are for example more workers than jobs, some workers will not be assigned a job.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%image?

@@

The resulting function can then be used to analyze from which distributions the data-points may have been drawn, which may also prove useful when analyzing clustering solutions.

@@

At first, the user tried sampling k-Means with $k$ in the range $2$ to $20$ creating $10$ solutions each. Doing this, he looked at which values for $k$ resulted in outlying/non-robust results. When filtering for solutions with $\leq10$ clusters, it was easy to see that the results with more than $10$ clusters produce poor solutions. The effect of which clusterings are filtered out this way can be seen in Figure \ref{fig:user_qcm_optics}.

@@

To start of, the user tried creating clustering solutions using DBSCAN, but in both the reachability plot and the heat map it could be seen that no robust solutions except nearly trivial ones could be found. In most cases the points were either nearly all assigned to one cluster or different small clusters each. With this information he deducted, that the algorithm is probably not fit for finding a good solution and chose to use k-Means. There, as he did not know how many clusters are in the data-set, he sampled $k$ from $2$ to $50$ and found again that solutions with $k>20$ seemed poor. He decided on this value, just like in the first example, by finding where solutions begin to differ strongly. The resulting reachability plot is shown in Figure \ref{fig:user_wifi_optics}
